Liveness-Based Garbage Collection for Lazy Languages

### 2015-04-27 paper/paper.tex
Added Results and Conclusion section from previous paper

%% Finally, we use $\Lfonly$  to range over {\em demand transformers}.
%% These transform  demands on  a function call  into demands  for its
%% formal  parameters:   given  a   user  function  $f$,   defined  by
%% $(\DEFINE\ (f\  x_1\ \ldots\ x_n)\  \ e_f)$ and called  with demand
%% $\sigma$,  then $\Lf{f}{i}{\sigma}$  is  the liveness  of $x_i$  at
%% $e_f$.

@@

%%   Analogously to classical liveness being computed as a solution of
%% dataflow   equations,  we   require,  via   inference   rule  ({\sc
%% live-define}), $\Lfonly$ to  satisfy the fixed-point property that:
%% when we  assume $\Lfonly$ to  be the family of  demand transformers
%% for the program then the  calculated liveness of each function body
%% $\mathcal{L}(e_f,\sigma,\Lfonly)$    agrees   with    the   assumed
%% $\Lfonly$.   As  usual,  there  are  often  multiple  solutions  to
%% $\Lfonly$; all are  safe (see Section~\ref{sec:correctness}) but we
%% prefer  the least  one  as  giving the  least  liveness subject  to
%% safety---and hence greatest amount of garbage collected.

We next  describe the function $\mathcal{L}$  that propagates liveness
across  expressions.   This brings  us  to  the  second way  in  which
liveness analysis for lazy languages differ from eager languages---the
point  where  an  expression   gets  evaluated  cannot  be  statically
determined.  To  deal with this, we introduce an analysis which is
flow insensitive but path-sensitive\comment{Check the terminology} into
our  analysis.

\begin{figure*}[t!]
%%\begin{picture}(100,130)(0,-90)
\begin{picture}(100,130)(0, -50)
  \begin{tabular}{cc}
    \begin{boxedminipage}{.51\textwidth}
      {\sf
        \small
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\psi_1\!\!:\,$ \px
            \UNL{4} \hspace*{1.15cm} $\psi_2\!\!:\, (\RETURN ~0)$
            \UNL{4} \hspace*{1.15cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{1.15cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{1.15cm} $\pi_7\!\!:\,
            $(\LET~\pz\ $\leftarrow$ (1~+~\py))))))) \IN
            \UNL{7} \hspace*{1.15cm} $\psi_3\!\!:\, (\RETURN ~\pz)$)))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $\pi_\mainpgm\!\!:\, \pi_8\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
	  \UNL{2}   \hspace*{.78cm}   $\pi_9\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
          \UNL{3}   \hspace*{.78cm}    $\pi_{10}\!\!:\,
          $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN
          \UNL{4}  \hspace*{.78cm}  $\psi_{4}\!\!:\,$ (\RETURN\ \pw)))))))
\end{uprogram}}
        } 
    \end{boxedminipage}
    & 
    \begin{boxedminipage}{.49\textwidth}
      {\sf
        \small
        \begin{tabular}{lc}
        $\cal{M_{\mbox{\tiny length}}} = $ 
            \begin{tabular}{lc}            
              $[$
              $ep_2 \mapsto \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \
              \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace ,$ \\ 
              $ep_1 \mapsto \lbrace \rbrace , $\\
              $ep_3 \mapsto \lbrace \px.\epsilon, \xl.\epsilon \rbrace $  \\
              $]$
            \end{tabular} \\
            $\cal{P_{\mbox{\tiny length}}} = [$
              \begin{tabular}{lc}
                $\pi_{\mbox{1}} \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \
                $ \pi_{\mbox{2}} \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \\
                $ \pi_{\mbox{3}} \mapsto \lbrace ep_3 \rbrace, $ \
                $ \pi_{\mbox{4}} \mapsto \lbrace ep_1 \rbrace, $ \
                $ \pi_{\mbox{5}} \mapsto \lbrace ep_2 \rbrace, $ \\
                $ \pi_{\mbox{6}} \mapsto \lbrace ep_2 \rbrace, $ \
                $ \pi_{\mbox{7}} \mapsto \lbrace ep_2 \rbrace $ \\
                $]$
              \end{tabular}\\
              $\cal{M_{\mbox{\tiny main}}} = $ 
              \begin{tabular}{lc}            
                $[$
                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace $ \
                  $]$
              \end{tabular} \\
              $\cal{P_{\mbox{\tiny main}}} = [$
                \begin{tabular}{lc}
                  $\pi_{\mbox{8}} \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \pi_{\mbox{9}} \mapsto \lbrace ep_4 \rbrace, $ \\
                  $ \pi_{\mbox{10}} \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \pi_{\mbox{11}} \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \ldots  $ \
                  $]$  
                \end{tabular}
        \end{tabular}
      }
    \end{boxedminipage}
    \\
    (a) Example program.&
    \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
        (b) {\em e-path} to liveness environment map \\
        (c) Prog pt to {\em e-paths } map
    \end{tabular}}
  \end{tabular}
\end{picture}
\kern -3ex
\vspace{0.5in}
\caption{Example Program and its liveness maps.}\label{fig:mot-example2}    
%% >>>>>>> 67d5a8ebe45703d9726af91b752e9d74152ae455
\end{figure*}

@@

Section~\ref{sec:computing}   shows   how   the  demand   transformers
$\Lfonly$  for  a  program  (representing  a  fully  context-sensitive
analysis) can  be safely  approximated, for each  function, by  a {\em
  procedure summary} (unifying the  contexts in the style of 0-CFA)\@.
The  summary consists  of a  pair  of a  single demand  and, for  this
demand, the corresponding  tuple of demands the function  makes on its
arguments.
\section{Computing liveness and its encoding as a table}\label{sec:computing}

Section~\ref{sec:liveness} gave a context-sensitive liveness
analysis and {\color {red}  proved it correct with reference
  to   a  {\em   minefield}   semantics}.   In   particular,
\mmap{f}{\sigma}{\Lfonly}  (and  the  map \pmap{f}  together
describes  a  liveness  set  for  each program  point  in  a
function in terms of a  given $\sigma$ and \Lfonly. We still
have not mentioned what is the specific sigma that should be
used for each function How is LF to be computed.

@@

\section{Results}
%Add graphs for memort usage LGC vs RGC
%Add graphs/table showing the GC times and other statistics
\section{Conclusion and Future Work}
%Mention lower peak memory usage. 
%Mention faster GC than full LGC and more garbage collected than RGC.
%Pitch it as sweet spot between full LGC and RGC.
%Future work
%Mention full LGC
%Mention k-liveness, use length example to show 1-liveness is not sufficient.
%Mention mixed mode GC, 
% - doing RGC most of the times and switching to LGC in case RGC fails to collect any garbage
% - doing LGC most of the times but regularly checking the reachability of the heap and 
% switching to RGC if LGC almost equals RGC

>>>>

@@

We next  describe the function $\mathcal{L}$  that propagates liveness
across  expressions.   This brings  us  to  the  second way  in  which
liveness analysis for lazy languages differ from eager languages---the
point  where  an  expression   gets  evaluated  cannot  be  statically
determined.  To  deal with this, we introduce an analysis which is
flow-insensitive but path-sensitive\comment{Check the terminology} into
our  analysis.

\begin{figure*}[t!]
\begin{picture}(100,130)(0, -50)
  \begin{tabular}{cc}
    \begin{boxedminipage}{.51\textwidth}
      {\sf
        \small
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\psi_1\!\!:\,$ \px
            \UNL{4} \hspace*{1.15cm} $\psi_2\!\!:\, (\RETURN ~0)$
            \UNL{4} \hspace*{1.15cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{1.15cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{1.15cm} $\pi_7\!\!:\,
            $(\LET~\pz\ $\leftarrow$ (1~+~\py))))))) \IN
            \UNL{7} \hspace*{1.15cm} $\psi_3\!\!:\, (\RETURN ~\pz)$)))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $\pi_\mainpgm\!\!:\, \pi_8\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
	  \UNL{2}   \hspace*{.78cm}   $\pi_9\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
          \UNL{3}   \hspace*{.78cm}    $\pi_{10}\!\!:\,
          $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN
          \UNL{4}  \hspace*{.78cm}  $\psi_{4}\!\!:\,$ (\RETURN\ \pw)))))))
\end{uprogram}}
        } 
    \end{boxedminipage}
    & 
    \begin{boxedminipage}{.49\textwidth}
      {\sf
        \small
        \begin{tabular}{lc}
        $\cal{M_{\mbox{\tiny \length}}} = $ 
            \begin{tabular}{lc}            
              $[$
              $ep_2 \mapsto \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \
              \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace ,$ \\ 
              $ep_1 \mapsto \lbrace \rbrace , $\\
              $ep_3 \mapsto \lbrace \px.\epsilon, \xl.\epsilon \rbrace $  \\
              $]$
            \end{tabular} \\
            $\cal{P_{\mbox{\tiny \length}}} = [$
              \begin{tabular}{lc}
                $\pi_{\mbox{1}} \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \
                $ \pi_{\mbox{2}} \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \\
                $ \pi_{\mbox{3}} \mapsto \lbrace ep_3 \rbrace, $ \
                $ \pi_{\mbox{4}} \mapsto \lbrace ep_1 \rbrace, $ \
                $ \pi_{\mbox{5}} \mapsto \lbrace ep_2 \rbrace, $ \\
                $ \pi_{\mbox{6}} \mapsto \lbrace ep_2 \rbrace, $ \
                $ \pi_{\mbox{7}} \mapsto \lbrace ep_2 \rbrace $ \\
                $]$
              \end{tabular}\\
              $\cal{M_{\mbox{\tiny \main}}} = $ 
              \begin{tabular}{lc}            
                $[$
                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace $ \
                  $]$
              \end{tabular} \\
              $\cal{P_{\mbox{\tiny \main}}} = [$
                \begin{tabular}{lc}
                  $\pi_{\mbox{8}} \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \pi_{\mbox{9}} \mapsto \lbrace ep_4 \rbrace, $ \\
                  $ \pi_{\mbox{10}} \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \pi_{\mbox{11}} \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \ldots  $ \
                  $]$  
                \end{tabular}
        \end{tabular}
      }
    \end{boxedminipage}
    \\
    (a) Example program.&
    \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
        (b) {\em e-path} to liveness environment map \\
        (c) Prog pt to {\em e-paths } map
    \end{tabular}}
  \end{tabular}
\end{picture}
\kern -3ex
\vspace{0.5in}
\caption{Example Program and its liveness maps.}\label{fig:mot-example2}    
\end{figure*}

@@

Section~\ref{sec:computing}   shows   how   the  demand   transformers
$\Lfonly$  for  a  program  (representing  a  fully  context-sensitive
analysis) can  be safely  approximated, for each  function, by  a {\em
  procedure summary} (unifying the  contexts in the style of 0-CFA)\@.
The  summary consists  of a  pair  of a  single demand  and, for  this
demand, the corresponding  tuple of demands the function  makes on its
arguments.

\section{Computing liveness and its encoding as a table}\label{sec:computing}
Section~\ref{sec:liveness} gave a context-sensitive liveness
analysis and {\color {red}  proved it correct with reference
  to   a  {\em   minefield}   semantics}.   In   particular,
\mmap{f}{\sigma}{\Lfonly}  (and  the  map \pmap{f}  together
describes  a  liveness  set  for  each program  point  in  a
function in terms of a  given $\sigma$ and \Lfonly. We still
have not mentioned what is the specific sigma that should be
used for each function How is LF to be computed.

@@

\section{Results}
To demonstrate the effectiveness of liveness-based garbage collection,
we  have  built a  prototype  consisting  of  an interpreter  for  our
language,  a  liveness  analyzer  and  a copying  collector  that  can
optionally use the results of liveness analysis for marking instead of
reachability. When the collector uses liveness for marking, we call it
a  {\em liveness-based  collector} (LGC),  else we  use the  term {\em
  reachability-based  collector}  (RGC).   The  copying  collector  is
neither incremental nor generational.  As a consequence, any cell that
becomes unreachable or  dead is assuredly collected in  the next round
of garbage collection.

While exploring the activation record of a function, LGC uses the DFAs
at  the latest  program  point traversed  in the function's body.  Thus, if  LGC
is invoked  at the program point   $\pi$,  it would  use  the
liveness  DFAs at  $\pi$ itself for exploring the root set of the
current activation record. For any other activation record in the
stack, say for a function ${\mathit f}$ calling ${\mathit g}$ in the call-chain,
the DFAs at the point of call to  ${\mathit g}$ would be used.

Let $\dfa{\pi}{x}$  denote the DFA  for the variable  and program
point  pair ($x$,  $\pi$).  We  assume that  there is  a function
$\mathsf{initial}(\dfa{\pi}{x})$  returning the initial  state of
$\dfa{\pi}{x}$.    
Considering   a   DFA   as   a   table,
$\dfa{\pi}{x}\mathit{(q,sym)}$  returns the  next  state for  the
state $q$ and the  symbol $\mathit{sym}$, where $\mathit{sym}$ is
\acar\      or     \acdr.       We      shall     also      write
$\dfa{\pi}{x}\mathit{(q,sym)\mathsf{?}}$    for    a    predicate
indicating  whether there  is a  transition from  $q$ on
$\mathit{sym}$.  The LGC action to chase the root variable $x$ at
$\pi$ can be described as  follows: If $\mathscr{L}(\dfa{\pi}{x})$ is
empty,  then  nothing  needs  to  be  done.   Otherwise  we  call
$\mathsf{copy}(\dfa{\pi}{x},\mathsf{initial}(\dfa{\pi}{x}),    x)$
in   Figure~\ref{algo:LGC-main-loop}  and  assign   the  returned
pointer to $x$.
The function $\mathsf{move\_to\_tospace}(x)$  copies the value of
$x$  in the  other semi-space  and returns  the new  address.  It
hides  details such as  returning the  forwarding pointer  if the
value  of $x$  is  already copied,  and  creating the  forwarding
pointer otherwise.

The graphs  in Figure~\ref{fig:memory-usage} show the  number of cells
in the heap over time for RGC and LGC---here time is measured in terms
of the number of cons cells allocated. In addition, they also show the
number of  reachable cells and the  number of cells  that are actually
live\footnote{This   is  statically   approximated  by   our  liveness
  analysis}.  Since  the programs have  different memory requirements,
we have tuned the size of heap for each program to ensure a reasonable
number of  collections. An invocation  of RGC decreases the  number of
cells  in heap  till  it touches  the  curve of  reachable cells.   An
invocation of LGC decreases the number  of heap cells to no lower than
the curve of live cells.

To construct the  reachable and live curves, we  record for every cell
its creation time (\CreateTime), its last use time (\UseTime), and the
earliest time  when the  cell becomes unreachable  and can  be garbage
collected (\CollTime).  For accurate recording of  \CollTime, we force
frequent invocations  of a reachability based collector  in a separate
run.     A    cell   is    live    at    time   \mbox{\exectime}    if
\mbox{$\mbox{\CreateTime}       \leq       \mbox{\exectime}       \leq
  \mbox{\UseTime}$}.       If      \mbox{$\mbox{\CreateTime}      \leq
  \mbox{\exectime} \leq \mbox{\CollTime}$}, it is reachable.

The benchmark programs are drawn from the {\tt no-fib} suite and other
sources and have been  manually converted to  ANF.  All  graphs  except  {\tt  fibheap}  show  fewer  garbage
collector  invocations for  LGC. The  number  of reachable  cells in  {\tt
  fibheap} increases  initially, comes  down before the  heap becomes gets
full and then  continues to remains low.  This results in  a single garbage
collection in both RGC and LGC.  The graphs also show the precision of
our liveness analysis.  For  all programs  except  {\tt nperms} and
{\tt lambda} (Figure~\ref{fig:memory-usage-lambda}),  LGC
manages to collect a good portion of the data that are not live.

\subsection{Results}
\begin{figure}[t] 
\centering
\begin{boxedminipage}{.75\textwidth}
{\bf function} $\mathsf{copy}$($\mathsf{dfa}$, $q$, $x$)\\ 
\noindent \hspace*{.35cm}$y \leftarrow \mathsf{move\_to\_tospace}(x)$\\
\noindent \hspace*{.35cm}{\bf if} $x$.{\bf tag} $\neq$ \CONS\ {\bf
  then} {\bf skip}\\ 
\noindent \hspace*{.25cm}  {\bf else} {\bf if}  $\mathsf{dfa}(q,0)\mathsf{?}$ {\bf then}   $y.\CAR = \mathsf{copy}(\mathsf{dfa}, \mathsf{dfa}(q,0),x.\CAR)$\\ 
\noindent \hspace*{0.95cm} {\bf if}  $\mathsf{dfa}(q, 1)\mathsf{?}$ {\bf then}  $y.\CDR = \mathsf{copy}(\mathsf{dfa}, \mathsf{dfa}(q,1),x.\CDR)$\\
\noindent \hspace*{.35cm}{\bf return} $y$
\end{boxedminipage}
\caption{Function       for       copying       a       root       set
  variable.}\label{algo:LGC-main-loop} \figrule

\end{figure}

\begin{figure}[p]
\begin{tabular}{@{}c@{}c@{}}
\hskip -4mm{\epsfig{file=sudoku.eps, height=4cm, width=6.10cm}}
&
{\epsfig{file=lcss.eps, height=4cm, width=6.10cm}}
\\
\hskip -4mm{\epsfig{file=gc_bench.eps, height=4cm, width=6.10cm}}
 &
{\epsfig{file=nperm.eps, height=4cm, width=6.10cm}}
\\
\hskip -4mm{\epsfig{file=fibheap.eps, height=4cm, width=6.10cm}}
&
{\epsfig{file=knightstour.eps, height=4cm, width=6.10cm}}
\\
\hskip -4mm{\epsfig{file=nqueens.eps, height=4cm, width=6.10cm}}
&
{\epsfig{file=treejoin.eps, height=4cm, width=6.10cm}}
\end{tabular}
 \caption{Memory  usage of  programs. The blue and the red curves
  indicate the number of cons cells in the active semi-space for RGC and
  LGC respectively. The black curve represents the number of reachable
  cells and the grey curve represents the number of cells that are
  actually live (of which liveness analysis does  a static
  approximation). x-axis is the time measured in number of cons-cells
  allocated. y-axis is the number of cons-cells.}
\label{fig:memory-usage} \figrule
\end{figure}
The  increased effectiveness  of LGC  over RGC  is also  shown  in the
tables  in  Figure~\ref{fig:experimental-results}.   The  first  table
provides  statistics  regarding the  analysis  itself.  The number  of
states and the analysis times  are within tolerable limits. Precision
of analysis refers to the percentage of dead cells that is collected by
LGC, averaged over all invocations.
The second table shows garbage  collection statistics for RGC and LGC.
LGC  collects larger garbage  per invocation,  drags cells  for lesser
time and requires a smaller heap size ({\em MinHeap}) for program to run
in comparison with  RGC.

There are a couple of  issues of concern.  The garbage collection time
is larger in the case of LGC for some programs. The reason is that the
cost of consulting the liveness DFA may outweigh the combined benefits
of  fewer   garbage  collections   and  fewer  markings   per  garbage
collection.   The  other issue  is  illustrated  by  the program  {\tt
  lambda}    (memory   usage  shown   separately    in
Figure~\ref{fig:memory-usage-lambda}).  As can be seen from the table
in Figure~\ref{fig:experimental-results}, the  number of
touched cells\footnote{These  are the  cells which are  visited during
  marking phase, often  more than once due to  sharing.} in this
example is much higher for
LGC.  This  increase is due to  excessive sharing among  heap nodes in
this program.  Note  that a node re-visited because  of sharing is not
explored  any   further  during  a  RGC   collection.   However,  this
curtailment cannot happen in  LGC because of the possibility that the
node,  re-visited in a  different liveness  state, may  mark a  set of
cells different from the earlier visit.
\begin{figure}[t]
\footnotesize

\centering

\scalebox{.9}{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
\hline
{Program} & {\sf sudoku} & {\sf lcss} & {\sf gc\_bench} &  {\sf
  nperm} &  {\sf fibheap} & {\sf knightstour} &
{\sf treejoin} & {\sf nqueens} & {\sf lambda}\\
\hline
\hline
{Time (msec)}&106.56 &2.19 &0.32 &1.16 &2.4 &3.05 &2.61 &0.71&8.68 \\
{DFA size} &4206 &726 &258 &526 &675 &922 &737 &241 & 694\\
{Precision(\%)} &98.4&98.8&99.9&87.1&100&94.3&99.6&98.8&79.7\\
\hline
\end{tabular}}

\centerline{(a)}

\medskip

\scalebox{0.87}{\begin{tabular}{| c | r | r | r | r | r | r  |  r | r | r | r | r | r | r |}
\hline
    & \multicolumn{2}{c|}{\# Collected} & \multicolumn{2}{c|}{\# Touched}
  & \multicolumn{2}{c|}{} & \multicolumn{2}{c|}{} &
\multicolumn{2}{c|}{} & \multicolumn{2}{c|}{GC time} \\

                            &   \multicolumn{2}{c|}{cells per GC}  
                            &   \multicolumn{2}{c|}{cells per GC} 
                            &   \multicolumn{2}{c|}{\#GCs} 
                            &   \multicolumn{2}{c|}{MinHeap} 
                            &   \multicolumn{2}{c|}{Avg. Drag} 
                            &   \multicolumn{2}{c|}{(sec)} \\
\cline{2-13}
{Program}    &
RGC & LGC & RGC & LGC  & RGC & LGC  &   RGC & LGC & RGC & LGC & RGC & LGC \\
\hline
\hline
    {\sf   sudoku}  &381 &1107 &1268 &281 &8 &3 & 1346  &338 &527 &5 &.007 & .034 \\
    {\sf  lcss } & 46522 &51101 &6216 &1363&8&7& 52301  &1701 &5147 &588 &.045 & .144 \\
    {\sf   gc\_bench}  &129179 &131067 &1894 &4&9&9& 131071   &6 &16970  &4 &.086 & .075 \\
     {\sf  nperm}  & 47586  &174478 &201585 &60882&14&4& 202597  &37507 &171878 &76618 &1.406 & .9  \\
    {\sf  fibheap} &249502  &251525 &5555 &2997&1&1& 254520  &13558 &78720 &0 &.006 & .014  \\
    {\sf  knightstour}  &2593 &314564 &907502 &319299&1161&10&508225   &307092 &206729 &82112 &464.902 & 14.124  \\
    {\sf  treejoin} & 288666  &519943 &297570 &5547&2&1& 525488  &7150 &212653 &1954 &.356 & .217 \\
    {\sf   nqueens} & 283822 &1423226 &2133001 &584143&46&9& 1819579  &501093 &521826 &39465 &70.314 & 24.811 \\     
    {\sf   lambda}  &205 & 556 &2072&90345 &23 &8&966 & 721  &303 &95 &.093 &2.49  \\ 
%    {\sf   fft }& &  & & &&&& ???? &???? &????? &????? &???? & ????  \\
%    {\sf  nboyer} & &  & &   & &  & & &  \\
%    {\sf   circsim?} & &  & &   & & & & &   \\
 \hline
\end{tabular}}

\centerline{(b)}
\vskip -2mm
\caption{Experimental results comparing RGC and LGC. Table (a) gives
  data related to liveness analysis, and (b) gives garbage collection
  data. The average drag time is measured in allocated cons cells.}
\label{fig:experimental-results}
\normalsize
\end{figure}

\section{Collecting more garbage can never slow things down}
\label{sec:lgc-always-better}
Since garbage collection is  effectively asynchronous to the allocator
thread, one might  worry as to how {\em  robust} our measurements are.
For example,  while LGC would,  in general, collect more  garbage than
RGC in the same heap state,  is it possible that on some programs, LGC
might do  a larger number  of collections? We  prove a result  to show
that   this  cannot   happen.    This  result   applies  to   classical
mark-and-sweep and copying garbage  collectors and, we hypothesize, to
generational collectors too.

\begin{lemma}
For the  same mutator,  a liveness-based collector  can never  do more
garbage collections than a reachability-based collector.
\end{lemma}

\begin{proof}Assume, as before,  that time is measured in terms of the number of
  cons cells allocated. Let us run  two copies of the mutator with the
  two    different   collectors    but    with   synchronous    memory
  allocations.   The   garbage    collections,   however,   happen
  asynchronously.

  To prove the lemma, it is enough to show the truth of the following 
statement: After
 every  LGC invocation,  the count of LGC invocations   is  no greater
than  RGC invocations.   The  base case  holds  since the  first
invocations of both GCs happen at the same time.  Assume the statement
to be true after $n$ invocations of LGC.  Since LGC copies a subset of
reachable cells, its heap would contain no more cells than RGC heap at
the end  of the $n$th invocation.   Thus either RGC  is invoked next before
LGC, or LGC and RGC are both invoked next at the same time.  In either
case, the statement holds after $n+1$ invocations of LGC.
\end{proof}

%% \begin{figure}[h]
%% \centerline{\epsfig{file=memory-usage.eps, width=12cm, height=15cm}}
%% \caption{Memory  usage of  programs. The  solid  lines represents
%%   number of  live cells and the  dashed lines show  the number of
%%   reachable cells.  The red and blue  lines represent invocations
%%   of RGC and  LGC---their heights indicate  the amount of memory
%%   copied.   {\color{red}{Caution:  Figures  are from  a  previous
%%       paper and are being used as placeholder}}}
%% \label{fig:memory-usage} \figrule
%% \end{figure}
%%\clearpage

\section{Related Work} Previous attempts to  increase  the space efficiency
of functional programs by additional  reclamation of memory fall in two
broad categories. In the first,  the program itself is instrumented to
manage  reclamation and reallocation  without the  aid of  the garbage
collector.    Such    attempts   include:   sharing    analysis   based
reallocation~\cite{jones89compile},                       deforestation
techniques~\cite{wadler88deforest,gill93ashort,chitil99deforest},   
methods  based  on   linear  logic~\cite{hofmann00linear}  and  region
analysis~\cite{tofte98region}.   Closer  to  our approach,  there  are
methods   that  enable   the   garbage  collector   to  collect   more
garbage~\cite{inoue88analysis,lee05static}   by
explicitly  nullifying  pointers  that  are not  live.   However,  the
nullification,  done   at  compile  time,   requires  sharing  (alias)
analysis.  Our  method, in contrast, does not  require sharing because
of the availability of the heap  itself at runtime.  To the best of our
knowledge, this is the first  attempt at liveness-based marking of the
heap during garbage collection.

\begin{figure}[t]
\centerline{\epsfig{file=lambda.eps, height=4cm, width=7cm}}
 \caption{Memory  usage and garbage collection pattern of  {\tt lambda}.} 
\label{fig:memory-usage-lambda} \figrule
\end{figure}

\section{Conclusions}
\label{sec:conclusion}
We have defined a notion of liveness on structured data;
this generalizes classical liveness and strong liveness.
We started with a general fully context-sensitive analysis which
we proved correct with respect to a minefield semantics
(this models the effect of garbage collection between every
evaluation step).

To avoid scalability issues (and to avoid performing part of
the liveness computation at run time)
we defined an 0-CFA version of this liveness analysis in which
demands for function $f$ at all calling contexts are conflated into 
a single demand $\sigma_f$.
This enabled us to treat the liveness equations symbolically obtaining
context-free grammars for liveness at each GC point (calls to user
functions and to $\CONS$).
These were then converted to DFAs for run-time consultation by the
garbage collector. Experiments confirm the precision of the analysis.

To obtain performance figures we compared a reachability-based garbage
collector with a liveness-based  collector.  This showed a decrease in
the  number   of  GCs,  more  garbage  collected   per  invocation.  A
significant benefit of LGC is that programs can run in smaller memory when
compared to RGC. This is potentially useful in situations where memory
is limited---as with embedded systems. For a majority of programs, the
garbage collection times were reduced.

One issue we  highlighted was that while fewer  nodes were marked (and
more garbage collected), sometimes  cons cells could be visited and
traversed  multiple times  with different  sets of  liveness  paths to
explore.  Further work includes  improvements to the classical copying
collector to reduce the cost of this.

%Add graphs for memort usage LGC vs RGC
%Add graphs/table showing the GC times and other statistics

%Mention lower peak memory usage. 
%Mention faster GC than full LGC and more garbage collected than RGC.
%Pitch it as sweet spot between full LGC and RGC.
%Future work
%Mention full LGC
%Mention k-liveness, use length example to show 1-liveness is not sufficient.
%Mention mixed mode GC, 
% - doing RGC most of the times and switching to LGC in case RGC fails to collect any garbage
% - doing LGC most of the times but regularly checking the reachability of the heap and 
% switching to RGC if LGC almost equals RGC
